Contracts for Contracts

### 2023-06-08 paper/contract_solidity.tex
sec 2

We now introduce \lang informally with examples.
\lang is a non-breaking extension that additionally provides means
to specify behavioral contracts on top of the Solidity language.
We provide a brief introduction of Solidity first, then discuss our extension
of behavioral contracts.
We often use \emph{specification} for behavioral contracts to avoid terminological ambiguity.

\paragraph{Design Rationale}
This paper concerns about designing an extension for Solidity that enhances the
language with higher-order temporal behavioral contracts.
The extension should be
\todo{non-intrusive, easy to use, expressive, gradual (you do not need to pay the overhead
	if you don't use contracts), efficient (little overhead)}

@@

Specifying the pre-condition and post-condition, where the post-condition
can depend on the argument value:

% \begin{lstlisting}[language=Solidity]
%   # {x | x > 0}
%   # {y | y > x + 1}
%   function f(uint x) returns (uint y) { ... }
% \end{lstlisting}

\begin{lstlisting}[language=Solidity]
  # {x | x % 1e18 < 1e17}
  # {y | y < type(uint).max / 1e18}
  function toEther(uint x) returns (uint y) { ... }
\end{lstlisting}

It is also equivalent to core syntax:
% \begin{lstlisting}
%   # {x | x > 0} -> {y | y > x + 1}
% \end{lstlisting}

\begin{lstlisting}[language=Solidity]
  # {x | x % 1e18 < 1e17} -> {y | y < type(uint).max / 1e18}
  function toEther(uint x) returns (uint y) { ... }
\end{lstlisting}

We can omit any part of the argument or return value contracts:

% \begin{lstlisting}[language=Solidity]
%   # {x | x > 0}
%   function f(uint x) returns (uint y) { ... }
% \end{lstlisting}

\begin{lstlisting}[language=Solidity]
  # {x | x < type(uint).max / 1e18}
  function toWei(uint x) returns (uint y) { ... }
\end{lstlisting}

It is equivalent to core syntax where the omitted part is simply \code{true}:

\begin{lstlisting}[language=Solidity]
  # {x | x < type(uint).max / 1e18} -> {y | true}
\end{lstlisting}

Another way is to use the `any` flat contract, which is defined as
\code{any = { _ | true }}:
\begin{lstlisting}
  # {x | x < type(uint).max / 1e18} -> any
\end{lstlisting}

For multi-argument functions, we use multiple \code{->} to compose larger function
contracts:

\begin{lstlisting}
  # {x | x > 0} -> {y | y < 100} -> {z | z == x + y }
  function f(int x, int y) returns (int z) { returns x + y; }
\end{lstlisting}

% \begin{lstlisting}[language=Solidity]
% # {rate | rate > 0 && rate < 100}
% # {time | time != 0}
% function simpleInterest(uint principle, uint rate, uint time) returns (uint interest) { ... }
% \end{lstlisting}

Or, we could write a single flat contract for the argument introducing
multiple argument binders:

\begin{lstlisting}
  # { (x, y) | x > 0 && y < 100 } -> {z | z == x + y }
  function f(int x, int y) returns (int z) { returns x + y; }
\end{lstlisting}

When specifying the spec of functions, two special variables are available
`msg.sender` the address of the caller and `block.timestamp` the current
block timestamp.
\zz{How about other special variables, such as `block.chainid` and `tx.origin`?. Perhaps we could redesign a universal syntax for these variables in Section 2.5, where we discuss where to add the message sender}
\url{https://docs.soliditylang.org/en/develop/units-and-global-variables.html#block-and-transaction-properties}

\begin{lstlisting}
  # { n | msg.value >= 1e15 * n }
  function buyTickets(int n) payable { ... }
\end{lstlisting}

>>>>

We now introduce \lang informally with examples.
\lang is a non-breaking extension that additionally provides means
to specify behavioral contracts on top of the Solidity language.
We often use \emph{specification} for behavioral contracts to avoid terminological ambiguity.

\paragraph{Design Rationale}
This paper concerns about designing an extension for Solidity that enhances the
language with contracts for higher-order temporal behaviors.
The extension should be
\begin{itemize}
  \item non-intrusive
  \item expressive
  \item gradual (you do not need to pay the overhead if you don't use contracts)
  \item efficient (no or little overhead)
\end{itemize}

@@

We begin with specifying pre-condition and post-condition for functions
involving first-order values. Consider the following \code{toEither} example
that converts \note{Wei?} to Either:
\begin{lstlisting}[language=Solidity]
# { toEither(x) returns (y)
    requires x % 1e18 < 1e17
    ensures  y < type(uint).max / 1e18 }
function toEther(uint x) returns (uint y) { ... }
\end{lstlisting}

The first line introduces bindings for \code{toEither}'s arguments and return
values, which are \code{x} and \code{y}, respectively.
The binding names in the specification do not have to match those in the function definition.
The \code{requires}-clause specifies the pre-condition to call \code{toEither},
and the \code{ensures}-clause specifies the post-condition.
Any occurrences of \code{x} in the \code{requires}-clause refer to the actual
argument value, and similarly, occurrences of \code{y} in the \code{ensures}-clause
refer to the actual returned value.

\paragraph{Omitting}
We can omit any part of the argument or return value specification:
\begin{lstlisting}
# { toWei(x) returns (_)
    ensures x < type(uint).max / 1e18 }
function toWei(uint x) returns (uint) { ... }
\end{lstlisting}
which is equivalent to core form where the omitted part is simply the boolean expression
\code{true}:
\begin{lstlisting}[language=Solidity]
# { toWei(x) returns (_)
    ensures x < type(uint).max / 1e18
    requires true }
\end{lstlisting}

\paragraph{Dependent Contract}
Note that post-condition can depend on the argument as well.
\todo{example}

%Another way is to use the `any` flat contract, which is defined as
%\code{any = { _ | true }}:
%\begin{lstlisting}
%  # {x | x < type(uint).max / 1e18} -> any
%\end{lstlisting}

\paragraph{Any Expression is Allowed}
When specifying the pre-/post-conditions of functions, the programmer
can use any valid Solidity expression to from the condition.
For example, Solidity has a number of built-in special variables
carrying important data (e.g. \code{msg.value}) or metadata (e.g. \code{block.timestamp})
of the transacation.
The programmer may want to check conditions against with those transactional
information.
The following code snippet shows an example that examines
\code{msg.value} as the pre-condition of a payable function:
\begin{lstlisting}
# { byTickets(n) ensures msg.value >= 1e15 * n }
function buyTickets(int n) payable { ... }
\end{lstlisting}
%\zz{How about other special variables, such as `block.chainid` and `tx.origin`?. Perhaps we could redesign a universal syntax for these variables in Section 2.5, where we discuss where to add the message sender}
%\url{https://docs.soliditylang.org/en/develop/units-and-global-variables.html#block-and-transaction-properties}

So far, boring.
